---
title: "02: Description and imputation"
execute: 
  warning: false
toc: true
toc-depth: 3
toc-expand: true
fig-width: 16
fig-asp: 0.618  # golden ratio
format:
  html:
    grid:
      body-width: 950px
    embed-resources: true
# knitr:
#   opts_chunk:
#     out.width: "100%"
---

Here we will explore our variables, and impute the datasets if necessary.

# Setup

```{r}
library(tidyverse)
theme_set(theme_bw(base_size = 16))

library(scales)     # plot utilities
library(ggokabeito) # colorblind palette
library(ggh4x)      # facet_nested
library(patchwork)  # arrange plots
library(glue)       # string interpolation
library(gt)         # make tables

library(mgcv)       # GAMs
library(gratia)     # utility functions for GAMs
library(marginaleffects)  # model interpretation
```


```{r}
base_path <- ".."
data_path <- file.path(base_path, "data")
res_path <- file.path(base_path, "results")


# data_path <- "local_data_path"
```


# Population

```{r}
# Load
pop <- read_csv(file.path(data_path, "processed", "population.csv"))


# Visualize population
pop |>
  ggplot() +
  geom_vline(
    aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
  ) +
  geom_line(
    aes(x=date, y=population, color=district), linewidth=0.8
  ) +
  geom_point(
    aes(x=date, y=population, color=district), size=3
  ) +
  scale_y_continuous(labels = scales::comma, n.breaks = 6) +
  scale_x_date(
    date_breaks="year", date_labels = "%Y",
    # limits = date_range_plot
  ) +
  scale_color_okabe_ito() +
  theme_bw(base_size = 18) +
  labs(
    color = "District",
    x = "Year",
    y = "Population"
  )

ggsave(
  file.path(res_path, "plots", "population.png"),
  width=16, height=16*0.600
)
```

Population has been increasing linearly year by year. Imputing the population of 2022 with a linear prediction seems appropriate.

```{r}
# Get time index for regression
pop_reg <- pop |> 
  mutate(year=year(date))

# Fit
fit <- lm(population ~ year*district, data=pop_reg)
summary(fit)

# Predict
newdata <- data.frame(
  district = c("Moshi", "Siha"),
  year = rep(2022, 2),
  date = rep(as.Date("2022-01-01"), 2)
)

newdata <- cbind(
  newdata,
  population = round(predict(fit, type="response", newdata=newdata))
)

newdata

# Join
pop_imp <- pop_reg |> 
  full_join(newdata) |> 
  select(- year) |> 
  arrange(district, date)


# Check imputation
pop_imp |>
  left_join(mutate(pop, population_orig = population)) |> 
  mutate(
    imputed = if_else(is.na(population_orig), "Yes", "No")
  ) |> 
  ggplot() +
  geom_vline(
    aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
  ) +
  geom_line(
    aes(x=date, y=population, color=district), linewidth=0.8
  ) +
  geom_point(
    aes(x=date, y=population, color=district, shape=imputed, size=imputed),
  ) +
  scale_shape_manual(
    values = c(16,4)
  ) +
  scale_size_manual(
    values = c(3,8)
  ) +
  scale_y_continuous(labels = scales::comma, n.breaks = 6) +
  scale_x_date(
    date_breaks="year", date_labels = "%Y",
  ) +
  scale_color_okabe_ito() +
  theme_bw(base_size = 18) +
  labs(
    color = "District",
    x = "Year",
    y = "Population",
    shape = "Imputed",
    size = "Imputed"
  )

ggsave(
  file.path(res_path, "plots", "population_imputed.png"),
  width=16, height=16*0.600
)

# Save
write_csv(
  pop_imp,
  file.path(data_path, "processed", "population_imputed.csv")
)


rm(fit, newdata, pop_reg)
```


# Environment

TODO: visualize splines

We will fit GAMs to decompose the environmental time series into trend and seasonal components. These models will be used to impute observations in the year range between 2014-2021, to make possible the posterior regression analyses.


```{r}
df_environ <- read_csv(file.path(data_path, "processed", "environmental.csv"))

# Prepare data for modelling
df_environ_model <- df_environ |> 
  mutate(
    time_ind = interval(min(date), date) / months(1) + 1,
    month_ind = month(date),
    district = factor(district, levels=c("Moshi", "Siha"))
  )

df_environ_model |> 
  print(width=Inf, n=13)
```


```{r}
df_environ_model

# Create dataframe for analysis
# Add necesary parameters to do analysis for each environmental variable

df_environ_an <- tribble(
  ~var, ~label, ~ units,
  "pm2p5", "PM2.5", "(\U03BCg/m\U00B3)",
  "greenness", "Greenness", "(NDVI)",
  "temp_min", "Min. temp.", "(ºC)",
  "temp_mean", "Mean temp.", "(ºC)",
  "temp_max", "Max. temp.", "(ºC)",
  "utci", "UTCI", "",
  "total_rainfall", "Rainfall", "(mm)",
  "n_raindays", "No. rain days", ""
)

df_environ_an
```


```{r}

# Step 1: Plot the response --------------------------------------------------------------

# Get range for plots
get_range <- function(var) {
  df_environ_model |> 
    filter(!is.na( .data[[var]] )) |> 
    pull(date) |> 
    range()
}

df_environ_an <- df_environ_an |> 
  mutate(
    date_range_plot = map(var, get_range)
  )

df_environ_an[["date_range_plot"]]


# Create plots
plots <- pmap(
  df_environ_an,
  function(var, label, units, date_range_plot, ...) {
    df_environ_model |>
      ggplot() +
      geom_vline(
        aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
      ) +
      geom_line(
        aes(x=date, y=.data[[var]], color=district), linewidth=0.8
      ) +
      geom_point(
        aes(x=date, y=.data[[var]], color=district), size=2.5
      ) +
      scale_x_date(
        date_breaks="year", date_labels = "%Y",
        limits = date_range_plot
      ) +
      scale_color_okabe_ito() +
      theme_bw(base_size = 18) +
      labs(
        color = "District",
        x = "Date (month)",
        y = glue("{label} {units}")
      )
  }
)

paths <- file.path(res_path, "plots", "environmental_descriptive",
                   glue("env_{df_environ_an$var}_response.png"))
# Save plots
walk2(
  plots,
  paths,
  \(plot, path) ggsave(
    filename=path, plot=plot,
    width=16, height=16*0.600
  )
)
```



```{r}
# Step 2: Fit the model ----------------------------------------------------------------

# Penalized GAM
# Contains penalized smooth functions of time and month
# Function of time (trend) is cubic spline with 1 basis function per year
# Function of month (seasonal component) is cyclic cubic splines with 12 basis functions
# We allow the seasonal and trend components to interact, so the seasonal component can
# change over time. The interaction is computed with a tensor product of the two functions

df_environ_an

# Get number of basis functions for trend
get_k_trend <- function(var) {
  df_environ_model |> 
    filter(! is.na(.data[[var]])) |>
    pull(date) |>
    year() |>
    range() |>
    diff()+1
}

df_environ_an <- df_environ_an |> 
  mutate(
    k_trend = map_dbl(var, get_k_trend)
  )

df_environ_an[["k_trend"]]

# Add distribution family of the response
df_environ_an <- df_environ_an |> 
  mutate(
    family = c(
      replicate(6, gaussian(), simplify=FALSE),
      list(
        mgcv::tw(link="log"),
        mgcv::nb(link="log")
      )
    )
  )
df_environ_an[["family"]]

# Fit models
fit_descriptive_gam <- function(var, k_trend, family, ...) {
  # Formula
  form <- as.formula(glue("{var} ~ 0"))
  form <- update(
    form,
    . ~ 0 + district +
      ti(time_ind, bs="cr", k=k_trend, by=district) +
      ti(month_ind, bs="cc", k=12, by=district) +
      ti(time_ind, month_ind, bs=c("cr", "cc"), k=c(k_trend,12), by=district)
  )
  
  print(form)
  
  # Start and end points of monthly cyclic splines
  # note: Slightly out of c(1,12) to not impose that January (1) is equal to December (12)
  knots <- list(month_ind = c(0.5, 12.5))
  
  gam(
    form,
    data=df_environ_model,
    family = family,
    method = "REML",
    knots = knots
  )
}

df_environ_an <- df_environ_an |> 
  mutate(
    model = pmap(list(var, k_trend, family), fit_descriptive_gam)
  )
```


```{r}
# Step 3: Check the model --------------------------------------------------------------

pwalk(
  df_environ_an,
  function(var, model, ...) {
    cat(glue(
      "\n
      ---------------------------------------
      Printing model diagnostics of variable:
                       {var}
      ----------------------------------------
      \n"
    ))
    
    print(summary(model))
    
    # Residual diagnostics
    p_diag <- gratia::appraise(model)
    # Save plot
    ggsave(
      plot=p_diag,
      filename = file.path(
        res_path, "plots", "environmental_descriptive",
        glue("env_{var}_gam_diagnostics.png")
      ),
      width=16, height=16*0.618
    )
    
    # Optionally: check model component predictions
    # gratia::draw(model) |> print()
  }
)
```


```{r}
# Step 4: Get predictions, as well as decomposition into trend, seasonal and residual ----

# Get predictions and time series decomposition
get_pred_ts_decomp <- function(model, model_df=df_environ_model, var, ...) {
  # note: Standard errors computed with Delta method (marginaleffects default)
  
  # Dataframe of predictions
  pred_df <- model_df
  
  # Select columns of interest
  pred_df <- pred_df |> 
    select(district, date, all_of(var), time_ind)
  
  # Get predictions for the response
  pred_name <- "pred"
  pred_df_sub <- predictions(
    model,
    newdata = model_df,
    type = "response"
  ) |> 
    as_tibble() |> 
    select(time_ind, district, estimate, conf.low, conf.high) |> 
    rename(
      "{pred_name}" := estimate,
      "{pred_name}_conf.low" := conf.low,
      "{pred_name}_conf.high" := conf.high,
    )
  pred_df <- pred_df |> 
    left_join(pred_df_sub)
  
  # Predictions for trend
  pred_name <- "trend"
  exclude_set <- c(
    as.character(glue("ti(month_ind):district{levels(pred_df$district)}")),
    as.character(glue("ti(time_ind,month_ind):district{levels(pred_df$district)}"))
  )
  pred_df_sub <- predictions(
    model,
    newdata = model_df,
    exclude = exclude_set,
    type = "response"
  ) |> 
    as_tibble() |> 
    select(time_ind, district, estimate, conf.low, conf.high) |> 
    rename(
      "{pred_name}" := estimate,
      "{pred_name}_conf.low" := conf.low,
      "{pred_name}_conf.high" := conf.high,
    )
  pred_df <- pred_df |> 
    left_join(pred_df_sub)
  
  # Predictions for seasonality
  pred_name <- "season"
  exclude_set <- c(
    "district",
    as.character(glue("ti(time_ind):district{levels(pred_df$district)}"))
  )
  pred_df_sub <- predictions(
    model,
    newdata = model_df,
    exclude = exclude_set,
    type = "response",
  ) |> 
    as_tibble() |> 
    select(time_ind, district, estimate, conf.low, conf.high) |> 
    rename(
      "{pred_name}" := estimate,
      "{pred_name}_conf.low" := conf.low,
      "{pred_name}_conf.high" := conf.high,
    )
  pred_df <- pred_df |> 
    left_join(pred_df_sub)
  
  # Add response residuals
  pred_df <- pred_df |> 
    left_join(cbind(model$model, residuals = residuals(model, type="response")))
  
  return(pred_df)
}

df_environ_an <- df_environ_an |> 
  mutate(
    pred_df = pmap(list(model=model, var=var), get_pred_ts_decomp)
  )
```


```{r}
# Step 5: Plot time series decomposition -----------------------------------------------

# df_environ_an$pred_df[[1]] |> colnames()

pwalk(
  df_environ_an,
  function(pred_df, var, date_range_plot, label, units, ...) {
    print(var)
    
    # Reference bar for y scale
    barlength <- select(pred_df, c(all_of(var), pred:season_conf.high)) |> 
      sapply(\(x) diff(range(x, na.rm=TRUE))) |>
      min()
    
    # Plot fitted line
    p_fit <- pred_df |> 
      filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
      ggplot() +
      geom_vline(
        aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
      ) +
      geom_vline(
        aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
      ) +
      geom_point(
        aes(x=date, y=.data[[var]], color=district), size=2.5
      ) + 
      geom_line(
        aes(x=date, y=pred, color=district), linewidth=0.8
      ) +
      geom_ribbon(
        aes(x=date, ymin=pred_conf.low, ymax=pred_conf.high, fill=district),
        alpha=0.2
      ) +
      geom_rect(
        aes(
          xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
          ymin=mean(.data[[var]], na.rm=TRUE) - (barlength/2),
          ymax=mean(.data[[var]], na.rm=TRUE) + (barlength/2)
        ),
        color="black", fill="gray75"
      ) +
      scale_x_date(
        date_breaks="year", date_labels = "%Y"
      ) +
      scale_color_okabe_ito() +
      scale_fill_okabe_ito() +
      theme_bw(base_size = 18) +
      labs(
        color = "District",
        fill = "District",
        x = "Date (month)",
        y = glue("{label} {units}")
      )
    
    # Plot trend
    p_trend <- pred_df |> 
      filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
      ggplot() +
      geom_vline(
        aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
      ) +
      geom_vline(
        aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
      ) +
      geom_line(
        aes(x=date, y=trend, color=district), linewidth=0.8
      ) +
      geom_ribbon(
        aes(x=date, ymin=trend_conf.low, ymax=trend_conf.high, fill=district),
        alpha=0.2
      ) +
      geom_rect(
        aes(
          xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
          ymin=mean(trend, na.rm=TRUE) - (barlength/2),
          ymax=mean(trend, na.rm=TRUE) + (barlength/2)
        ),
        color="black", fill="gray75"
      ) +
      scale_x_date(
        date_breaks = "year", date_labels = "%Y"
      ) +
      scale_color_okabe_ito() +
      scale_fill_okabe_ito() +
      theme_bw(base_size = 18) +
      labs(
        color = "District",
        fill = "District",
        x = "Date (month)",
        y = "Trend"
      )
    
    # Plot seasonal component
    p_season <- pred_df |> 
      filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
      ggplot() +
      geom_vline(
        aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
      ) +
      geom_vline(
        aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
      ) +
      geom_line(
        aes(x=date, y=season, color=district), linewidth=0.8
      ) +
      geom_ribbon(
        aes(x=date, ymin=season_conf.low, ymax=season_conf.high, fill=district),
        alpha=0.2
      ) +
      geom_rect(
        aes(
          xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
          ymin=mean(season, na.rm=TRUE) - (barlength/2),
          ymax=mean(season, na.rm=TRUE) + (barlength/2)
        ),
        color="black", fill="gray75"
      ) +
      scale_x_date(
        date_breaks = "year", date_labels = "%Y"
      ) +
      scale_color_okabe_ito() +
      scale_fill_okabe_ito() +
      theme_bw(base_size = 18) +
      labs(
        color = "District",
        fill = "District",
        x = "Date (month)",
        y = "Seasonality"
      )
    
    # Plot residuals
    # p_resid <- pred_df |>
    #   filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
    #   ggplot() +
    #   geom_vline(
    #     aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    #   ) +
    #   geom_vline(
    #     aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    #   ) +
    #   geom_segment(
    #     aes(x=date, y=0, yend=residuals, color=district), linewidth= 1.5, lineend = "butt",
    #     position = position_dodge(width=20)
    #   ) +
    #   geom_hline(yintercept = 0) +
    #   ylab("residuals") +
    #   geom_rect(
    #     aes(
    #       xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
    #       ymin=mean(residuals, na.rm=TRUE) - (barlength/2),
    #       ymax=mean(residuals, na.rm=TRUE) + (barlength/2)
    #     ),
    #     color="black", fill="gray75"
    #   ) +
    #   scale_x_date(
    #     date_breaks = "year", date_labels = "%Y"
    #   ) +
    #   scale_color_okabe_ito() +
    #   scale_fill_okabe_ito() +
    #   theme_bw(base_size = 18) +
    #   labs(
    #     color = "District",
    #     x = "Date (month)",
    #     y = "Residuals"
    #   )
    
    # Merge plots
    p_joint <- p_fit / (p_trend + theme(legend.position = "none")) /
      (p_season + theme(legend.position = "none"))  +
      # (p_resid + theme(legend.position = "none")) +
      plot_layout(axes = "collect", guides="collect",
                  heights = c(4,3,3))
    
    # Save plot
    ggsave(
      plot=p_joint,
      filename=file.path(
        res_path, "plots", "environmental_descriptive",
        glue("env_{var}_gam_decomposition.png")
      ),
      width=16, height=16*0.700
    )
  }
)
```



```{r}
# Step 6: Get imputations needed for the regression model --------------------------------
# Full observations in year range 2014-2022


# Dataframe to store imputations
df_environ_imp <- df_environ

# Dataframe with the predictions
df_environ_pred <- df_environ_an |> 
  unnest(pred_df) |> 
  select(var, district, date, pred) |> 
  pivot_wider(
    names_from = var,
    values_from = pred
  )


# Make sure both dataframes are in the same order and are of the same dimensions
df_environ_imp <- df_environ_imp |> 
  arrange(district, date)

df_environ_pred <- df_environ_pred |> 
  arrange(district, date)

dim(df_environ_imp)
dim(df_environ_pred)

# Replace NAs with predictions -----.
# For each variable
for (var in df_environ_an$var) {
  print(var)
  
  # Get indices where NA
  na_ind <- which(is.na(df_environ_imp[[var]]))
  # Substitute predictions
  df_environ_imp[na_ind, var] <- df_environ_pred[na_ind, var]
}

# Fix n_raindays so that it is integer
df_environ_imp <- mutate(df_environ_imp, n_raindays = round(n_raindays))

# Filter imputations in year range 2012-2021 (cut out 2022)
df_environ_imp <- filter(df_environ_imp, year(date) <= 2021)

# Plot imputations -----------------------------------------------------------------------

# Prepare df for plotting
df_environ_imp_plot <- df_environ_imp |> 
  left_join(df_environ, join_by(district, date), suffix=c(".imp", ".org"))

for (i in seq_len(nrow(df_environ_an))) {
  var <- df_environ_an[["var"]][i]
  
  # If no NA, skip
  vec <- df_environ_imp_plot[[glue("{var}.org")]]
  if(!any(is.na(vec))) next
  
  print(var)
  # Else, plot imputation
  df_environ_imp_plot |>
    mutate(
      imputed = if_else(is.na(.data[[glue("{var}.org")]]), "Yes", "No")
    ) |> 
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_line(
      aes(x=date, y=.data[[glue("{var}.imp")]], color=district), linewidth=0.8
    ) +
    geom_point(
      aes(x=date, y=.data[[glue("{var}.imp")]], color=district, shape=imputed, size=imputed),
    ) +
    scale_shape_manual(
      values = c(16,4)
    ) +
    scale_size_manual(
      values = c(2.5,8)
    ) +
    scale_x_date(
      date_breaks="year", date_labels = "%Y",
    ) +
    scale_color_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      x = "Date (month)",
      y = glue("{df_environ_an[['label']][i]} {df_environ_an[['units']][i]}"),
      shape = "Imputed",
      size = "Imputed"
    )
  
  ggsave(
    file.path(res_path, "plots", "environmental_descriptive",
              glue("env_{var}_response_imputed.png")),
    width=16, height=16*0.600
  )
}

# Save dataframe with imputed environmental variables
write_csv(
  df_environ_imp,
  file.path(data_path, "processed", "environmental_imputed.csv")
)
```



```{r}
# Step 7: Make contrasts for the trend -------------------------------------------------
# note: Standard errors and hypothesis tests computed with Delta method
# (marginaleffects default)

# Check time indices for contrast
# df_environ_model |>
#   filter(date %in% c("2012-01-01", "2021-12-01", "2022-12-01")) |>
#   print(width=Inf)

make_trend_contrasts <- function(model, var, date_range_plot, label, units, ...) {
  print(var)
  
  # Exclude terms to make predictions for the trend
  exclude_set <- c(
    as.character(glue("ti(month_ind):district{levels(df_environ_model$district)}")),
    as.character(glue("ti(time_ind,month_ind):district{levels(df_environ_model$district)}"))
  )
  
  # Average slope ----------------------------------------------------------------------
  
  # Average change in the trend per 1 year increase
  avg_yearly_trend_slope <- avg_comparisons(
    model,
    variables = list(time_ind = 12),
    by = "district",
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
  ) |> 
    as_tibble() |> 
    mutate(
      name = "avg_yearly_trend_slope"
    )
  
  avg_yearly_trend_slope
  
  
  # Difference in the slopes (Moshi - Siha)
  avg_yearly_trend_slope_diff <- avg_comparisons(
    model,
    variables = list(time_ind = 12),
    by = "district",
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
    # Test for the equality of both estimates
    hypothesis = "b2 - b1 = 0"
  ) |> 
    as_tibble() |> 
    mutate(
      name = "avg_yearly_trend_slope_diff"
    )
  
  avg_yearly_trend_slope_diff
  
  
  # Plot yearly trend change before taking the average by district ---.
  
  # plot_comparisons(
  #   model,
  #   variables = list(time_ind = 12),
  #   by = c("time_ind", "district"),
  #   comparison = "difference",
  #   type = "response",
  #   exclude = exclude_set,
  # )
  
  p_yearly_trend_slope <- comparisons(
    model,
    newdata = df_environ_model,
    variables = list(time_ind = 12),
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
  ) |> 
    as_tibble() |> 
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |>
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_hline(yintercept = 0) +
    geom_line(
      aes(x=date, y=estimate, color=district), linewidth=0.8
    ) +
    geom_ribbon(
      aes(x=date, ymin=conf.low, ymax=conf.high, fill=district),
      alpha=0.2
    ) +
    scale_x_date(
      date_breaks="year", date_labels = "%Y"
    ) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      fill = "District",
      x = "Date (month)",
      y = glue("{label} yearly trend change\n{units}")
    )
  
  
  # Average trend difference -----------------------------------------------------------
  
  # Average trends
  avg_trend <- avg_predictions(
    model,
    variables = "district",
    type = "response",
    exclude = exclude_set
  ) |> 
    as_tibble() |> 
    mutate(
      name = "avg_trend"
    )
  
  avg_trend
  
  # Average difference in trends
  avg_trend_diff <- avg_comparisons(
    model,
    variables = "district",
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
  ) |> 
    as_tibble() |> 
    mutate(
      name = "avg_trend_diff"
    )
  
  avg_trend_diff
  
  # Plot trend difference before taking the average by district ----.
  
  # plot_comparisons(
  #   model,
  #   variables = c("district"),
  #   by = c("time_ind"),
  #   comparison = "difference",
  #   type = "response",
  #   exclude = exclude_set,
  # )
  
  p_trend_diff <- comparisons(
    model,
    newdata = df_environ_model,
    variables = "district",
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
  ) |> 
    as_tibble() |> 
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |>
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_hline(yintercept = 0) +
    geom_line(
      aes(x=date, y=estimate), linewidth=0.8
    ) +
    geom_ribbon(
      aes(x=date, ymin=conf.low, ymax=conf.high),
      alpha=0.2
    ) +
    scale_x_date(
      date_breaks="year", date_labels = "%Y"
    ) +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      fill = "District",
      x = "Date (month)",
      y = glue("{label} trend difference\n(Siha - Moshi) {units}")
    )
  
  # Trend difference between 2012-01 and 2021-12 ---------------------------------------
  
  start_indices <- c(time_ind=1, month_ind=1)
  end_indices <- c(time_ind=120, month_ind=12)
  
  # Alternatively ---.
  # Each variable has observations available for different time ranges, so might want
  # to set different time indices
  # PM2.5 has observations until 2022-12
  # The rain variables for Siha rely on extrapolations for the contrast
  # ---.
  # start_indices <- df_environ_model |> 
  #   filter(date == date_range_plot[[1]]) |> 
  #   select(time_ind, month_ind) |> 
  #   distinct() |> 
  #   as_vector()
  # 
  # end_indices <- df_environ_model |> 
  #   filter(date == date_range_plot[[2]]) |> 
  #   select(time_ind, month_ind) |> 
  #   distinct() |> 
  #   as_vector()
  # -------------------.
  
  # Query explanation:
  # For the observations we had at 2012-01 (time_ind==1, month_ind==1),
  # what is the effect on the trend of changing time_ind to 120 and month_ind to 12 (date being 2021-12)
  
  # Temporal difference
  temp_trend_diff <- comparisons(
    model,
    newdata = filter(df_environ_model, time_ind==1),
    variables = list(
      time_ind = c(start_indices[["time_ind"]], end_indices[["time_ind"]]),
      month_ind = c(start_indices[["month_ind"]], end_indices[["month_ind"]])
    ),
    cross = TRUE,
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
  ) |> 
    as_tibble() |> 
    relocate(district) |> 
    mutate(
      name = "temp_trend_diff"
    )
  
  temp_trend_diff
  
  # Difference in temporal differences
  temp_trend_diff_diff <- comparisons(
    model,
    newdata = filter(df_environ_model, time_ind==1),
    variables = list(
      time_ind = c(start_indices[["time_ind"]], end_indices[["time_ind"]]),
      month_ind = c(start_indices[["month_ind"]], end_indices[["month_ind"]])
    ),
    cross = TRUE,
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
    # Test for the equality of both estimates
    hypothesis = "b2 - b1 = 0"
  ) |> 
    as_tibble() |> 
    mutate(
      name = "temp_trend_diff_diff"
    )
  
  temp_trend_diff_diff
  
  # Plot the trends
  p_trends <- predictions(
    model,
    newdata = df_environ_model,
    type = "response",
    exclude = exclude_set,
  ) |> 
    as_tibble() |> 
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |>
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      xintercept = as.Date(c("2012-01-01", "2021-12-01")), linewidth=1
    ) +
    geom_line(
      aes(x=date, y=estimate, color=district), linewidth=0.8
    ) +
    geom_ribbon(
      aes(x=date, ymin=conf.low, ymax=conf.high, fill=district),
      alpha=0.2
    ) +
    scale_x_date(
      date_breaks="year", date_labels = "%Y"
    ) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      fill = "District",
      x = "Date (month)",
      y = glue("{label} trend {units}")
    )
  
  # Merge results ----------------------------------------------------------------------
  
  # Combine earlier plots
  p_joint <- p_trends / p_yearly_trend_slope / p_trend_diff +
    plot_layout(axes="collect")
  
  ggsave(
    plot=p_joint,
    filename=file.path(
      res_path, "plots", "environmental_descriptive",
      glue("env_{var}_trend_contrasts.png")
    ),
    width=16, height=16*0.700
  )
  
  # Combine tables
  contrast_res <- bind_rows(
    avg_yearly_trend_slope,
    avg_yearly_trend_slope_diff,
    avg_trend,
    avg_trend_diff,
    temp_trend_diff,
    temp_trend_diff_diff
  ) |>
    select(! c(predicted_lo:predicted, date:time_ind)) |> 
    relocate(name)
  
  return(contrast_res)
}

df_environ_an <- df_environ_an |> 
  mutate(
    contrast_res = pmap(list(model, var, date_range_plot, label, units), make_trend_contrasts)
  )

df_environ_an[["contrast_res"]][[1]] |> print(width=Inf)
```


```{r}

# Prepare table for contrasts

contrast_tbl <- df_environ_an |> 
  select(var, label, units, contrast_res) |> 
  unnest(contrast_res) |> 
  select(label, name, district, estimate, conf.low, conf.high, p.value) |> 
  # pull(name) |> unique()
  mutate(
    Contrast = case_match(
      name,
      "avg_yearly_trend_slope" ~ "Average yearly trend change",
      "avg_yearly_trend_slope_diff" ~ "Average yearly trend change",
      "avg_trend" ~ "Average trend level",
      "avg_trend_diff" ~ "Average trend level",
      "temp_trend_diff" ~ "Trend difference (2012-01 to 2021-12)",
      "temp_trend_diff_diff" ~ "Trend difference (2012-01 to 2021-12)",
    ),
    District = if_else(is.na(district), "Siha - Moshi", district),
    `Estimate (95% CI)` = glue("{round(estimate, 3)} ({round(conf.low, 3)}, {round(conf.high, 3)})"),
    `p value` = Hmisc::format.pval(p.value, digits=3),
  ) |> 
  select(- c(estimate, conf.low, conf.high, p.value, district, name)) |> 
  gt(
    groupname_col = "label",
    rowname_col = "Contrast"
  )

contrast_tbl

# Save
contrast_tbl |> 
  gtsave(
    file.path(res_path, "tables", "environmental_contrasts.html")
  )

```


# Disease

## Prepare data

```{r}
df_disease <- read_csv(file.path(data_path, "processed", "disease.csv"))
```

```{r}
# Add population (2022 imputed) to disease dataframe
df_disease_model <- df_disease |> 
  mutate(year = year(date)) |> 
  left_join(
    mutate(pop_imp, year=year(date)),
    join_by(district, year)
  ) |> 
  select(- c(year, date.y)) |> 
  rename(date = date.x)

# Compute incidence rate per 100k people
df_disease_model <- df_disease_model |> 
  mutate(case_rate = n_cases / population * 1e5)


# Set up variables for modelling
df_disease_model <- df_disease_model |> 
  mutate(
    time_ind = interval(min(date), date) / months(1) + 1,
    month_ind = month(date),
    district = factor(district, levels=c("Moshi", "Siha"))
  )

df_disease_model |> 
  print(width=Inf, n=13)

```

```{r}
df_disease_model |> 
  filter(disease == "Malaria", district=="Siha") |> 
  arrange(date) |> 
  print(n=Inf)
```


```{r}

# Re-define disease groupings according to Renz and Skevaki notations
# Type manually

# Renz

ref_renz <- tribble(
  ~disease_group_renz, ~disease,
  # 01
  "Gastrointestinal Infections", c(
    "Diarrhea With No Dehydration",
    "Diarrhea With Severe Dehydration",
    "Dysentery",
    "Cholera",
    "Typhoid",
    "Intestinal Worms"
  ),
  # 02
  "Urinary Infections", c(
    "Urinary Tract Infections"
  ),
  # 03
  "Respiratory Infections", c(
    "Upper Respiratory Infections",
    "Tuberculosis",
    "Pneumonia, Severe",
    "Influenza"
  ),
  # 04
  "Other infections", c(
    "Meningitis",
    "Rabies",
    "Measles",
    "Leprosy",
    "Infectious Eye Disease",
    "Skin Infection - Fungal",
    "Malaria"
  ),
  # 05
  "Malnutrition", c(
    "Nutritional Disorders, Other",
    "Moderate Malnutrition",
    "Marasmus",
    "Marasmic Kwashiorkor",
    "Kwashiorkor"
  ),
  # 06
  "Diabetes Mellitus", c(
    "Diabetes Mellitus"
  ),
  # 07
  "Bronchial Asthma", c(
    "Bronchial Asthma"
  ),
  # 08
  "Trauma and Injuries", c(
    "Road Traffic Accidents",
    "Fractures"
  ),
  # 09
  "Cardiovascular Diseases", c(
    "Other Cardiovascular Diseases",
    "Hypertension"
  ),
  # 10
  "Cancer", c(
    "Neoplasms/Cancer"
  )
) |> 
  unnest(cols=c(disease))

ref_renz |> 
  print(n=Inf)


```


```{r}
# Skevaki

ref_skevaki_block <- tribble(
  ~disease_block_skevaki, ~disease_group_skevaki,
  # A
  "Infectious/Communicable Diseases", c(
    # a
    "Gastrointestinal Infections",
    # b
    "Vector-borne Infections",
    # c
    "Respiratory Infections",
    # d
    "Other Communicable Diseases"
  ),
  # B
  "Disorders of Mixed Etiology", c(
    # e
    "Diarrheal Diseases",
    # f
    "Neurological Diseases"
  ),
  # C
  "Non-Communicable Diseases", c(
    # g
    "Cardiovascular Diseases",
    # h
    "Respiratory Diseases",
    # i
    "Gynecological Diseases",
    # j
    "Diabetes Mellitus",
    # k
    "Malnutrition",
    # l
    "Cancer",
    # m
    "Neurological",
    # n
    "Trauma",
    # o
    "Other Non-Communicable"
  )
) |> 
  unnest(cols=c(disease_group_skevaki))

# ---------------------------------------------

ref_skevaki_group <- tribble(
  ~disease_group_skevaki, ~disease,
  # A
  # a
  "Gastrointestinal Infections", c(
    "Diarrhea With Severe Dehydration",
    "Dysentery",
    "Cholera",
    "Typhoid",
    "Intestinal Worms"
  ),
  # b
  "Vector-borne Infections", c(
    "Malaria"
  ),
  # c
  "Respiratory Infections", c(
    "Upper Respiratory Infections",
    "Tuberculosis",
    "Pneumonia, Severe",
    "Influenza"
  ),
  # d
  "Other Communicable Diseases", c(
    "Schistosomiasis",
    "Urinary Tract Infections",
    "Rabies",
    "Measles",
    "Leprosy",
    "Infectious Eye Disease",
    "Skin Infection - Fungal"
  ),
  # B
  # e
  "Diarrheal Diseases", c(
    "Diarrhea With No Dehydration"
  ),
  # f
  "Neurological Diseases", c(
    "Meningitis",
    "Acute Flaccid Paralysis"
  ),
  # C
  # g
  "Cardiovascular Diseases", c(
    "Other Cardiovascular Diseases",
    "Hypertension"
  ),
  # h
  "Respiratory Diseases", c(
    "Bronchial Asthma"
  ),
  # i
  "Gynecological Diseases", c(
    "Gynecological Diseases"
  ),
  # j
  "Diabetes Mellitus", c(
    "Diabetes Mellitus"
  ),
  # k
  "Malnutrition", c(
    "Nutritional Disorders, Other",
    "Moderate Malnutrition",
    "Marasmus",
    "Marasmic Kwashiorkor",
    "Kwashiorkor"
  ),
  # l
  "Cancer", c(
    "Neoplasms/Cancer"
  ),
  # m
  "Neurological", c(
    "Psychoses",
    "Neuroses",
    "Epilepsy"
  ),
  # n
  "Trauma", c(
    "Road Traffic Accidents",
    "Fractures"
  ),
  # o
  "Other Non-Communicable", c(
    "Substance Abuse",
    "Snake and Insect Bites",
    "Poisoning",
    "Peptic Ulcers",
    "Caries"
  )
) |>
  unnest(cols=c(disease))


ref_skevaki <- ref_skevaki_block |> 
  full_join(ref_skevaki_group)

ref_skevaki |> 
  print(n=Inf)

rm(ref_skevaki_block, ref_skevaki_group)
```


```{r}
# Add new disease grouping to model table

df_disease_model <- df_disease_model |> 
  left_join(ref_renz) |> 
  left_join(ref_skevaki)
```

## Explore

Explore disease incidence rates and the different disease groupings.


```{r}
# Original disease classification

df_disease_model |> 
  # More space
  mutate(
    disease_group = recode(
      disease_group,
      `Diarrheal disease/Gastrointesntinal infections`="Diarrheal disease/\nGastrointesntinal infections",
      `Other Communicable Diseases/Gastrointestinal infections`="Other Communicable Diseases/\nGastrointestinal infections"
    )
  ) |>
  ggplot() +
  geom_tile(
    aes(x=date, y=disease, fill=case_rate),
    color="black",
    width=31
  ) +
  scale_fill_viridis_c(
    transform = "log1p",
    breaks = c(0,10,100,1000,6000),
  ) +
  facet_nested(
    cols = vars(district),
    rows = vars(disease_communicable, disease_group),
    scales = "free_y",
    space = "free_y",
    # strip = strip_vanilla(size = "variable", clip = "off"),
    strip = strip_nested(size = "variable", clip="off", bleed = FALSE),
    solo_line = TRUE,
    nest_line = TRUE,
    # switch = "y"
  ) +
  guides(
    fill = guide_colorbar(
      barheight = 12,
      title.vjust = 2,
    )
  ) +
  labs(
    fill = "Incidence rate\n(per 100k people)",
    x = "Date (month)",
    y = ""
  ) +
  theme(
    legend.title = element_text(face="bold"),
    axis.title = element_text(face="bold"),
    strip.text = element_text(face="bold"),
    # strip.background = element_rect(fill="white", color="gray"),
    strip.background = element_blank(),
    ggh4x.facet.nestline = element_line(colour = "black"),
    strip.text.y = element_text(angle=0, size=rel(1)),
    strip.text.x = element_text(angle=0, size=rel(1.5)),
    legend.text = element_text(size=rel(1.0))
  )


ggsave(
  file.path(res_path, "plots", "disease_incidence-rate_original-groups.png"),
  width=22, height=22*0.530
)

```


```{r}
# Original groups again, but remove category with many missings

df_disease_model |> 
  # More space
  mutate(
    disease_group = recode(
      disease_group,
      `Diarrheal disease/Gastrointesntinal infections`="Diarrheal disease/\nGastrointesntinal infections",
      `Other Communicable Diseases/Gastrointestinal infections`="Other Communicable Diseases/\nGastrointestinal infections"
    )
  ) |>
  # Remove disease category with many missings
  filter(disease_group != "Other vector borne diseases") |> 
  ggplot() +
  geom_tile(
    aes(x=date, y=disease, fill=case_rate),
    color="black",
    width=31
  ) +
  scale_fill_viridis_c(
    transform = "log1p",
    breaks = c(0,10,100,1000,6000),
  ) +
  facet_nested(
    cols = vars(district),
    rows = vars(disease_communicable, disease_group),
    scales = "free_y",
    space = "free_y",
    # strip = strip_vanilla(size = "variable", clip = "off"),
    strip = strip_nested(size = "variable", clip="off", bleed = FALSE),
    solo_line = TRUE,
    nest_line = TRUE,
    # switch = "y"
  ) +
  guides(
    fill = guide_colorbar(
      barheight = 12,
      title.vjust = 2,
    )
  ) +
  labs(
    fill = "Incidence rate\n(per 100k people)",
    x = "Date (month)",
    y = ""
  ) +
  theme(
    legend.title = element_text(face="bold"),
    axis.title = element_text(face="bold"),
    strip.text = element_text(face="bold"),
    # strip.background = element_rect(fill="white", color="gray"),
    strip.background = element_blank(),
    ggh4x.facet.nestline = element_line(colour = "black"),
    strip.text.y = element_text(angle=0, size=rel(1)),
    strip.text.x = element_text(angle=0, size=rel(1.5)),
    legend.text = element_text(size=rel(1.0))
  )


ggsave(
  file.path(res_path, "plots", "disease_incidence-rate_original-groups_minus-missings.png"),
  width=22, height=22*0.530
)

```


```{r}

# Skevaki groups

df_disease_model |> 
  filter(
    !is.na(disease_group_skevaki)
  ) |> 
  ggplot() +
  geom_tile(
    aes(x=date, y=disease, fill=case_rate),
    color="black",
    width=31
  ) +
  scale_fill_viridis_c(
    transform = "log1p",
    breaks = c(0,10,100,1000,6000),
  ) +
  facet_nested(
    cols = vars(district),
    rows = vars(disease_block_skevaki, disease_group_skevaki),
    scales = "free_y",
    space = "free_y",
    strip = strip_nested(size = "variable", clip="off", bleed = FALSE),
    solo_line = TRUE,
    nest_line = TRUE,
  ) +
  guides(
    fill = guide_colorbar(
      barheight = 12,
      title.vjust = 2,
    )
  ) +
  labs(
    fill = "Incidence rate\n(per 100k people)",
    x = "Date (month)",
    y = ""
  ) +
  theme(
    legend.title = element_text(face="bold"),
    axis.title = element_text(face="bold"),
    strip.text = element_text(face="bold"),
    # strip.background = element_rect(fill="white", color="gray"),
    strip.background = element_blank(),
    ggh4x.facet.nestline = element_line(colour = "black"),
    strip.text.y = element_text(angle=0, size=rel(1)),
    strip.text.x = element_text(angle=0, size=rel(1.5)),
    legend.text = element_text(size=rel(1.0))
  )


ggsave(
  file.path(res_path, "plots", "disease_incidence-rate_skevaki-groups.png"),
  width=22, height=22*0.530
)

```

```{r}
# Renz groups

df_disease_model |> 
  filter(
    !is.na(disease_group_renz)
  ) |>
  ggplot() +
  geom_tile(
    aes(x=date, y=disease, fill=case_rate),
    color="black",
    width=31
  ) +
  scale_fill_viridis_c(
    transform = "log1p",
    breaks = c(0,10,100,1000,6000),
  ) +
  facet_nested(
    cols = vars(district),
    rows = vars(disease_block_skevaki, disease_group_renz),
    # rows = vars(disease_group_renz),
    scales = "free_y",
    space = "free_y",
    strip = strip_nested(size = "variable", clip="off", bleed = FALSE),
    solo_line = TRUE,
    nest_line = TRUE,
  ) +
  guides(
    fill = guide_colorbar(
      barheight = 12,
      title.vjust = 2,
    )
  ) +
  labs(
    fill = "Incidence rate\n(per 100k people)",
    x = "Date (month)",
    y = ""
  ) +
  theme(
    legend.title = element_text(face="bold"),
    axis.title = element_text(face="bold"),
    strip.text = element_text(face="bold"),
    # strip.background = element_rect(fill="white", color="gray"),
    strip.background = element_blank(),
    ggh4x.facet.nestline = element_line(colour = "black"),
    strip.text.y = element_text(angle=0, size=rel(1)),
    strip.text.x = element_text(angle=0, size=rel(1.5)),
    legend.text = element_text(size=rel(1.0))
  )


ggsave(
  file.path(res_path, "plots", "disease_incidence-rate_renz-groups.png"),
  width=22, height=22*0.530
)

```


## Model trend and seasonality


```{r}
disease_analysis <- function(disease, disease_label, df, family){
  #' This function does descriptive analyses for each disease incidence rate time series
  #' 
  #' Arguments:
  #'  disease: str, disease name
  #'  df: dataframe with the data
  #'  family: distribution of the response
  #'  
  #'  Steps:
  #'    1: Plot the response
  #'    2: Fit the model
  #'    3: Validate model (plot diagnostics)
  #'    4: Compute predictions
  #'    5: Plot time series decomposition
  #'    6: Compute contrasts for the trend
  #'  
  #'  Returns:
  #'    A list with the model, predictions, plots and contrasts  
  
  
  cat(glue("
  \n\n\n
  ========================================================================================    
                        Doing analysis for disease variable:
                                        {disease}
  ========================================================================================
  
  "))
  
  
  # Filter dataframe for variable of interest
  
  df <- filter(df, disease == {{disease}})
  
  # Date range for which there are observations
  date_range_plot <- df |>
    filter(! is.na(n_cases)) |>
    pull(date) |>
    range()
  
  
  # Step 1: Plot the response ------------------------------------------------------------
  
  p_var <- df |>
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_line(
      aes(x=date, y=case_rate, color=district), linewidth=0.8
    ) +
    geom_point(
      aes(x=date, y=case_rate, color=district), size=2.5
    ) +
    scale_x_date(
      date_breaks="year", date_labels = "%Y",
      limits = date_range_plot
    ) +
    scale_color_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      x = "Date (month)",
      y = glue("{disease}\nincidence rate\n(per 100k people)")
    )
  
  ggsave(
    file.path(res_path, "plots", "disease_descriptive",
              glue("disease_{disease_label}_incidence-rate.png")),
    width=16, height=16*0.600
  )
  
  # Step 2: Fit the model ----------------------------------------------------------------
  
  # Penalized GAM
  # Contains penalized smooth functions of time and month
  # Function of time (trend) is cubic spline with 1 basis function per year
  # Function of month (seasonal component) is cyclic cubic splines with 12 basis functions
  # We allow the seasonal and trend components to interact, so the seasonal component can
  # change over time. The interaction is computed with a tensor product of the two functions
  
  # Number of basis functions for trend:
  k_trend <- df |>
    filter(! is.na(n_cases)) |>
    pull(date) |>
    year() |>
    range() |>
    diff()+1
  
  form <- n_cases ~ 0 + district +
    ti(time_ind, bs="cr", k=k_trend, by=district) +
    ti(month_ind, bs="cc", k=12, by=district) +
    ti(time_ind, month_ind, bs=c("cr", "cc"), k=c(k_trend,12), by=district)
  
  # Start and end points of monthly cyclic splines
  # note: Slightly out of c(1,12) to not impose that January (1) is equal to December (12)
  knots = list(month_ind = c(0.5, 12.5))
  
  mod <- gam(
    form,
    # Include the offset here instead of in the formula, so that it is not used in the predictions
    # (predictions are incidence rates per 100k people)
    offset = log(population/1e5),
    data=df,
    family = family,
    method = "REML",
    knots = knots
  )
  
  # Step 3: Check the model --------------------------------------------------------------
  
  # Check terms and significance
  print(summary(mod))
  # Check model assumptions
  p_diag <- gratia::appraise(mod)
  
  ggsave(
    plot=p_diag,
    file.path(res_path, "plots", "disease_descriptive",
              glue("disease_{disease_label}_gam_diagnostics.png")),
    width=16, height=16*0.618
  )
  
  # gratia::draw(mod)
  
  # Step 4: Get predictions, as well as decomposition into trend, seasonal and residual ----
  # note: Standard errors computed with Delta method (marginaleffects default)
  
  pred_df <- df
  
  # Get predictions for the response
  pred_name <- "pred"
  
  pred_df_sub <- predictions(
    mod,
    newdata = df,
    type = "response",
  ) |> 
    as_tibble() |> 
    select(time_ind, district, estimate, conf.low, conf.high) |> 
    rename(
      "{pred_name}" := estimate,
      "{pred_name}_conf.low" := conf.low,
      "{pred_name}_conf.high" := conf.high,
    )
  
  pred_df <- pred_df |> 
    left_join(pred_df_sub)
  
  # Predictions for trend
  pred_name <- "trend"
  
  exclude_set <- c(
    as.character(glue("ti(month_ind):district{levels(pred_df$district)}")),
    as.character(glue("ti(time_ind,month_ind):district{levels(pred_df$district)}"))
  )
  
  pred_df_sub <- predictions(
    mod,
    newdata = df,
    exclude = exclude_set,
    type = "response",
  ) |> 
    as_tibble() |> 
    select(time_ind, district, estimate, conf.low, conf.high) |> 
    rename(
      "{pred_name}" := estimate,
      "{pred_name}_conf.low" := conf.low,
      "{pred_name}_conf.high" := conf.high,
    )
  
  pred_df <- pred_df |> 
    left_join(pred_df_sub)
  
  # Predictions for seasonality
  pred_name <- "season"
  
  exclude_set <- c(
    "district",
    as.character(glue("ti(time_ind):district{levels(pred_df$district)}"))
  )
  
  pred_df_sub <- predictions(
    mod,
    newdata = df,
    exclude = exclude_set,
    type = "response",
  ) |> 
    as_tibble() |> 
    select(time_ind, district, estimate, conf.low, conf.high) |> 
    rename(
      "{pred_name}" := estimate,
      "{pred_name}_conf.low" := conf.low,
      "{pred_name}_conf.high" := conf.high,
    )
  
  pred_df <- pred_df |> 
    left_join(pred_df_sub)
  
  # Add response residuals
  pred_df <- pred_df |> 
    left_join(cbind(mod$model, residuals = residuals(mod, type="response")))
  
  pred_df |> print(width=Inf)
  
  # Step 5: Plot time series decomposition -----------------------------------------------
  
  # Reference bar for y scale
  barlength <- select(pred_df, c(case_rate, pred:season_conf.high)) |> 
    sapply(\(x) diff(range(x, na.rm=TRUE))) |> 
    min()
  
  # Plot fitted line
  p_fit <- pred_df |> 
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |>
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_point(
      aes(x=date, y=case_rate, color=district), size=2.5
    ) + 
    geom_line(
      aes(x=date, y=pred, color=district), linewidth=0.8
    ) +
    geom_ribbon(
      aes(x=date, ymin=pred_conf.low, ymax=pred_conf.high, fill=district),
      alpha=0.2
    ) +
    geom_rect(
      aes(
        xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
        ymin=mean(case_rate, na.rm=TRUE) - (barlength/2),
        ymax=mean(case_rate, na.rm=TRUE) + (barlength/2)
      ),
      color="black", fill="gray75"
    ) +
    scale_x_date(
      date_breaks="year", date_labels = "%Y"
    ) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    theme_bw(base_size = 16) +
    labs(
      color = "District",
      fill = "District",
      x = "Date (month)",
      y = glue("{disease}\nincidence rate\n(per 100k people)")
    )
  
  # Plot trend
  p_trend <- pred_df |> 
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_line(
      aes(x=date, y=trend, color=district), linewidth=0.8
    ) +
    geom_ribbon(
      aes(x=date, ymin=trend_conf.low, ymax=trend_conf.high, fill=district),
      alpha=0.2
    ) +
    geom_rect(
      aes(
        xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
        ymin=mean(trend, na.rm=TRUE) - (barlength/2),
        ymax=mean(trend, na.rm=TRUE) + (barlength/2)
      ),
      color="black", fill="gray75"
    ) +
    scale_x_date(
      date_breaks = "year", date_labels = "%Y"
    ) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      fill = "District",
      x = "Date (month)",
      y = "Trend"
    )
  
  # Plot seasonal component
  p_season <- pred_df |> 
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_line(
      aes(x=date, y=season, color=district), linewidth=0.8
    ) +
    geom_ribbon(
      aes(x=date, ymin=season_conf.low, ymax=season_conf.high, fill=district),
      alpha=0.2
    ) +
    geom_rect(
      aes(
        xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
        ymin=mean(season, na.rm=TRUE) - (barlength/2),
        ymax=mean(season, na.rm=TRUE) + (barlength/2)
      ),
      color="black", fill="gray75"
    ) +
    scale_x_date(
      date_breaks = "year", date_labels = "%Y"
    ) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      fill = "District",
      x = "Date (month)",
      y = "Seasonality"
    )
  
  # Plot residuals
  p_resid <- pred_df |>
    filter(date >= date_range_plot[1] & date <= date_range_plot[2]) |> 
    ggplot() +
    geom_vline(
      aes(xintercept = floor_date(date, "year")), linewidth=0.5, color="gray70"
    ) +
    geom_vline(
      aes(xintercept = max(floor_date(date, "year")) + years(1)), linewidth=0.5, color="gray70"
    ) +
    geom_segment(
      aes(x=date, y=0, yend=residuals, color=district), linewidth= 1.5, lineend = "butt",
      position = position_dodge(width=20)
    ) +
    geom_hline(yintercept = 0) +
    ylab("residuals") +
    geom_rect(
      aes(
        xmin=max(date_range_plot)+60, xmax=max(date_range_plot)+90,
        ymin=mean(residuals, na.rm=TRUE) - (barlength/2),
        ymax=mean(residuals, na.rm=TRUE) + (barlength/2)
      ),
      color="black", fill="gray75"
    ) +
    scale_x_date(
      date_breaks = "year", date_labels = "%Y"
    ) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    theme_bw(base_size = 18) +
    labs(
      color = "District",
      x = "Date (month)",
      y = "Residuals"
    )
  
  
  # Merge plots
  p_joint <- p_fit / (p_trend + theme(legend.position = "none")) /
    (p_season + theme(legend.position = "none"))  +
    # (p_resid + theme(legend.position = "none")) +
    plot_layout(axes = "collect", guides="collect")
  
  ggsave(
    plot=p_joint,
    file.path(res_path, "plots", "disease_descriptive",
              glue("disease_{disease_label}_gam_decomposition.png")),
    width=16, height=16*0.700
  )
  
  
  # Step 6: Make contrasts for the trend -------------------------------------------------
  # note: Standard errors computed with Delta method (marginaleffects default)
  
  # - Trend change between 2012-01 and 2021-12
  
  # -------------------------------------------.
  
  # - Trend change between 2012-01 and 2021-12
  
  # Check what values of time_ind and month_ind we need for the contrast
  # df |>
  #   filter(date %in% c("2012-01-01", "2021-12-01")) |>
  #   print(width=Inf)
  # time_ind=1, month_ind=1
  # time_ind=120, month_ind=12
  
  # Exclude terms to make predictions for the trend
  exclude_set <- c(
    as.character(glue("ti(month_ind):district{levels(pred_df$district)}")),
    as.character(glue("ti(time_ind,month_ind):district{levels(pred_df$district)}"))
  )
  
  cat("
    ------------------------------
    Doing contrasts for the trend:
    ------------------------------
    ")
  
  # Query explanation:
  # For the observations we had at 2012-01 (time_ind==1, month_ind==1),
  # what is the effect on the trend of changing time_ind to 120 and month_ind to 12 (date being 2021-12)
  
  # Difference
  contrast_trend_diff <- comparisons(
    mod,
    newdata = filter(df, time_ind==1),
    variables = list(time_ind = c(1, 120),
                     month_ind = c(1, 12)),
    cross = TRUE,
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
    
  ) |> 
    as_tibble()
  
  cat("
      ------------------------------------
      2012-01-01 to 2021-12-01 difference:
      
      ")
  contrast_trend_diff |> 
    print(width=Inf)
  
  # Difference in differences test
  contrast_trend_diff_diff <- comparisons(
    mod,
    newdata = filter(df, time_ind==1),
    variables = list(time_ind = c(1, 120),
                     month_ind = c(1, 12)),
    cross = TRUE,
    comparison = "difference",
    type = "response",
    exclude = exclude_set,
    # Test for the equality of both estimates
    hypothesis = "b1 = b2"
  ) |> 
    as_tibble()
  
  cat("
      ---------------------------------------------------
      2012-01-01 to 2021-12-01 difference in differences:
      
      ")
  contrast_trend_diff_diff |> 
    print(width=Inf)
  
  # Relative change
  contrast_trend_rc <- comparisons(
    mod,
    newdata = filter(df, time_ind==1),
    variables = list(time_ind = c(1, 120),
                     month_ind = c(1, 12)),
    cross = TRUE,
    comparison = "lift",
    type = "response",
    exclude = exclude_set
  ) |> 
    as_tibble()
  
  cat("
      -----------------------------------------
      2012-01-01 to 2021-12-01 relative change:
      
      ")
  contrast_trend_rc |> 
    print(width=Inf)
  
  # Difference in relative changes test
  contrast_trend_rc_diff <- comparisons(
    mod,
    newdata = filter(df, time_ind==1),
    variables = list(time_ind = c(1, 120),
                     month_ind = c(1, 12)),
    cross = TRUE,
    comparison = "lift",
    type = "response",
    exclude = exclude_set,
    hypothesis = "b1 = b2"
  ) |> 
    as_tibble()
  cat("
      -------------------------------------------------------
      2012-01-01 to 2021-12-01 difference in relative change:
      
      ")
  print(contrast_trend_rc_diff)
  
  # Results ------------------------------------------------------------------------------
  res <- list(
    mod = mod,
    pred_df = pred_df,
    plots = list(
      p_var = p_var,
      p_diag = p_diag,
      p_fit = p_fit,
      p_trend = p_trend,
      p_season = p_season,
      p_resid = p_resid,
      p_joint = p_joint
    ),
    contrasts = list(
      contrast_trend_diff = contrast_trend_diff,
      contrast_trend_diff_diff = contrast_trend_diff_diff,
      contrast_trend_rc = contrast_trend_rc,
      contrast_trend_rc_diff = contrast_trend_rc_diff
    )
  )
  
  return(res)
}

```

## Call

```{r}
# disease_names <- as.list(unique(df_disease_model$disease))
disease_names <- df_disease_model |> 
  filter(!is.na(disease_group_skevaki)) |> 
  pull(disease) |> 
  unique() |> 
  as.list()
names(disease_names) <- disease_names

call_list <- map(
  disease_names,
  \(x) x = list(disease=x,
                disease_label=str_replace_all(x, c(" "="-","/"="-")),
                df=df_disease_model,
                family=mgcv::nb(link="log"))
)

# call_list[["Diabetes Mellitus"]]
# disease="Diabetes Mellitus"
# disease_label=str_replace_all(disease, c(" "="-","/"="-"))
```


```{r}
res <- map(
  call_list,
  \(x) do.call(disease_analysis, x)
)

# Save results
saveRDS(res, file.path(res_path, "R_output", "disease_descriptive.Rds"))
```

